Unilateral Divorce and the Rise of Informal Cohabitation

### 2020-09-28 draft_sub_con.tex
latest

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.
\documentclass[12pt]{article}
\usepackage[round, sort , authoryear]{natbib}

@@

\renewcommand\theadalign{bc}
\renewcommand\theadgape{\Gape[4pt]}
\renewcommand\cellgape{\Gape[4pt]}

@@

%%% PAGE DIMENSIONS
\usepackage[margin=2.5 cm]{geometry}
\usepackage{blindtext} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage[utf8]{inputenc} 
\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{epstopdf}
\usepackage[hang]{footmisc}
\usepackage{lipsum}
\usepackage{setspace}
% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{pgfplots}
\pgfplotsset{compat=1.13}
\usepackage{caption}
\usepackage{threeparttablex}
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}
%%% PACKAGES
\usepackage{placeins}
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
% These packages are all incorporated in the memoir class to one degree or another...
\usepackage{amsmath}
\numberwithin{table}{section}
\usepackage{cases}
\usepackage{graphicx}
%
\usepackage{float}
\usepackage{authblk}
\usepackage{pgfplots}
\usepackage{pdfpages}
\linespread{1.5}
\setlength{\footnotemargin}{4mm}
\usepackage{amssymb} 
\usepackage{tabularx}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\addtolength{\footnotesep}{2mm} % change to 1mm

@@

%For subtables
\usepackage{subcaption}

%For having the catption above tables
\usepackage{float}
\floatstyle{plaintop}
\restylefloat{table}

%Math stuff
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\restylefloat{table}
\oddsidemargin  +0.18in
\evensidemargin +0.18in
\topmargin 40pt \textheight 8.1in \textwidth 6.5in
\linespread{1.5}\parskip .05in

%Nice Figure and table headers
\captionsetup[figure]{labelfont={sc},name={Figure},labelsep=period}
%\captionsetup[table]{labelfont={sc},name={Table},labelsep=period,justification=centering}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\parskip3mm\parindent0cm
\setlength{\droptitle}{-10em}   % This is your set screw

\title{Unilateral Divorce and the Rise of Informal Cohabitation\thanks{The authors wish to thank Matthias Doepke, Alessandra Voena and the Norhtwestern Macro group for their useful comments. Blasutto acknowledges financial support from the French speaking community of Belgium (ARC project 15/19-063 on "family transformations").}}
\author{Fabio Blasutto$^1$ \quad Egor Kozlov$^2$}

@@

	\footnotetext[1]{IRES, UCLouvain \& National Fund for Scientific Research (Belgium). Email:  \tt{fabio.blasutto@uclouvain.be}}
	\footnotetext[2]{Northwestern Univrsity. Email:  \tt{egorkozlov2020@u.northwestern.edu}.\\}

\begin{abstract}
	Does unilateral divorce eroded the gains of marriage with respect to informal cohabitation? Exploiting the staggered introduction of unilateral divorce across U.S. states, we show that after the reform newly formed relationships are more likely to be cohabitations instead of marriages, and that cohabitation spells last longer. To understand the mechanisms underlying the law changes, we build and estimate a structural life cycle model with partnership choice, where the gains from marriage with respect to cohabitation come from a better cooperation within the household, enforced through a costly divorce, which acts as a commitment technology. In the model the reform increase the risk of divorce, making cohabitation preferred to couples that would have had the highest risk of divorce. Since the couples switching relationship are on average better matched than the average cohabitants, the length of cohabitations increases.
\end{abstract}
\textbf{Keywords}: Marriage, Cohabitation, Unilateral Divorce, Structural Estimation\\
\textbf{JEL-Code}: D83 - J12
%\clearpage
%\noindent \medskip{}
\section{Introduction}
Cohabitation is on the rise: according to \cite{manning2013}, the share of women that ever cohabited moved from 33\% in 1987 to 60\% in 2010. Why did this happen?
Since the seminal work of Becker (\citealp{becker1981}), economists studied the incentives behind the decision to marry, among which the sharing of public goods, the division of labor to exploit comparative advantages (\citealp{chiappori1997}), and risk sharing (\citealp{voena2015} and \citealp{rigas2015}). These reasons tell us why couples live together, while they are actually silent about the choice between marriage and informal cohabitation, two partnership contracts subject to different rules. While the literature often simplifies, avoiding the distinction between these two, they display a different labor supply, wealth accumulation and educational composition. It is not surprising that households, the smallest economic unit, can behave differently when the contract ruling them is different, but this can be the key to understand important economic phenomena. For example, cohabitation display a much higher separation rate than marriages, which might be a cause of the high share of single mothers living in the US nowadays, which is the single strongest predictor for upward mobility\footnote{While the share of single mothers is the strongest predictor of differences in upward mobility across counties on top of segregation, residential segregation, income inequality, social capital, school quality, and racial shares, \cite{chetty2018} points out that half of this effect is due to self selection.} according to \cite{chetty2018}. If we want to understand the behavior of cohabiting households and the implication of it, we first need to understand their formation and why they are growing in number.

In this paper we address the question of the rise of cohabitation, focusing on the role of unilateral divorce on the share of people that decides to cohabit. Our main contribution is to show that the rate of cohabitation increased after the changes in the US law and that the underlying mechanism is that the increased risk of divorce makes marriage less attractive. As a consequence, cohabitation becomes the preferred choice to couples that would have experienced the highest risk of divorce. As a consequence, the average match quality of cohabiting couples increase, which increases the length of cohabitation spells.

We first use data from the National Survey of Family and the Household and of the National Survey of Family Growth to build a sample of first and second relationships\footnote{A relationship is defined as an interruption of the state of singleness, which can either be marriage or cohabitation.}. Then, exploiting the exogenous variation coming from the staggered introduction of unilateral divorce over time across US states, we show that cohabitation becomes around 5\% more likely  to be chosen compared to the pre-reform period. Interestingly,  the effect is heterogeneous on how property is divided upon divorce, being strongest in states where each spouse gets half of the wealth. This suggest that the result cannot be entirely due to a deterioration of divorce as a commitment technology\footnote{This mechanism is highlighted by \cite{lafortune2019}, where the equal division of assets within marriage makes possible to efficiently invest in children even after the rise of unilateral divorce.}, since the egalitarian property division rule can act as a substitute of the mutual consent regime. Moreover, we analyze how unilateral divorce affected the length of newly formed cohabitation spells, showing that they last longer both because of a reduced risk of marriage and separation.
To understand the mechanisms that lead to these changes we build a dynamic model of intrahousehold decision making, where cohabitation and marriage differ in their splitting cost as well as in the way property is divided. While in the case of separation assets are always split as an agreement between spouses, in the case of divorce in some states it is the judge to decide. Moreover, separation can be initiated unilaterally, as opposed to divorce, that would need the consensus of both partner under a mutual divorce regime. Individuals are initially single, and with some probability in each period they meet a potential partner, which they can decide to marry or start cohabiting with. Couples make decisions about consumption, savings and female labor supply and they are subject to idiosyncratic income shocks. Couples also receive time varying match quality shocks, which might drive the couple to change their partnership status (i.e. from cohabitation to marriage) or to separate.  We model the decision making in the couple building on the literature of limited commitment (see for example \cite{kk1996},  \cite{ligon2002}, \cite{marcet2019} and \cite{pavoni2018}), which has been applied to dynamic collective models in the household by \cite{voena2015}, \cite{mazzocco2007}, \cite{foerster2019} and \cite{lise2018} among others. In this framework, when one member of the couple wishes to split, the couple rebargains such that the binding member is made indifferent between separation and staying in a couple\footnote{For particularly bad draws in the match quality or productivity shock this might not be possible and hence the couple would split.}. The gains of marriage with respect to cohabitation come from a better risk sharing and a more efficient specialization in the production of a public good\footnote{We assume that a public good is produced with money an female time. When females stop working for producing such good their potential productivity in the labor market decreases.}, which arise from a better commitment within the couple, enforced by the threat of a costly divorce, which acts as a commitment device. On the other hand, couples that would face a high risk of divorce if married (i.e. because of a low match quality draw) are more likely to choose cohabitation instead, since it implies a lower cost of separation while allowing to enjoy the gains from being in a couple. In the model, a switch from mutual consent to unilateral divorce causes couples to start cohabiting more, since the higher risk of divorce makes the expected value of marriage with respect to cohabitation lower. The effect would be particularly pronounced under community property division of assets upon divorce, since the richest part of the couple  would risk to lose most of its wealth. This mechanism implies that couples that would have married under the old regime, are cohabiting instead. Since those have a higher match quality than the average cohabiting couples, this selection drives down the risk of separation for cohabiting couples, as observed in the data.

We then estimate the model using the simulated method of moments to learn about the size of our main mechanism. We use as targets an array of moments regarding the mating market and the differential female labor supply for cohabiting and married couples. We find that the model is able to reproduce the results from our empirical evidence, thus validating our mechanism. We then run a series of counterfactual experiments to gain intuition about forces that might have contributed to the rise of cohabitation in the last decades. [In particular, we quantify the role of the wage structure and technological progress in the household sector, which increases the opportunity cost of home production with female time and decreases the gains coming from labor market specialization.]

The contribution of this paper is threefold. First, we document and explain how divorce laws affected the choice between marriage and cohabitation. This adds to the existing literature that documented the effects of divorce laws on the rate of divorce (\citealp{friedberg1998}, \citealp{wolfers2006}), female labor supply (\citealp{stevenson2008}, \citealp{voena2015}), savings (\citealp{voena2015}), assortative mating (\citealp{reynoso2019}) and prostitution (\citealp{ciacci2017}) among the others. As in \cite{reynoso2019}, we study not only the effect of unilateral divorce on married couples, but also on partnership choice. While her focus is on who marries whom and concludes that unilateral divorced raised the share of singles, we abstract from modeling the mating market and we focus on partnership choice, which allow us to conclude that the rate of singleness declined less than previously thought, since some people that were believed not to be in a couple were actually cohabiting. Our paper build on \cite{voena2015}, that studies how unilateral divorce interacted with property rights upon divorce affected household behavior of married couples. We extend her work both considering cohabitation as an alternative relationship and in analyzing the effects of these laws on sorting.
Second, our paper expands the literature on cohabitation, showing how the risk of divorce is essential for understanding partnership choices. Different papers in the literature highlighted various gains of marriage with respect cohabitation: commitment (\citealp{matouschek2008}), labor specialization within the couple \cite{gemici2014}, learning about match quality \cite{brien2006}, the interaction between the cost of divorce and learning \cite{blasutto2020} and  investment in children \cite{lafortune2019}. A paper close to ours is \cite{lafortune2019}, who highlight the role of assets as a collateral that enforces commitment and hence allows for optimal investment in children. In their model, a switch from mutual consent to unilateral divorce causes people without wealth to start cohabiting because they are left without a commitment technology. While we think that their mechanism is important, it cannot be the whole story since we observe that the switch toward cohabitation was the weakest in title based state, while their model would predict the opposite. In this paper we will quantify how much of the shift is due to the erosion of marriage's collateral versus the role of the increased risk of divorce. We extend the work of \cite{gemici2014} and \cite{blasutto2020}, who develop collective dynamic models with limited commitment, introducing asset accumulation decisions, which are shown to play an important role for commitment and hence for partnership choice, especially when interacted with property rights upon divorce.\\
Third, our paper speaks to the literature studying the changes of the family in the last decades. Recent works have investigated various channels through which the structure of families might have changed over time. \cite{albanesi2016} show the role of improved maternal health on the rise of female labor force participation, \cite{greenwood2016} instead study the role that technology and the wage structure played in the rise in assortative mating, female labor force participation, the share of divorced and the increase in the share of singles. Our paper extends this work, studying the effect of wage structure, gender wage gap and changes in the home production technology on the rise of cohabitation. Through a series of counterfactual experiments we are able to inspected which were the main forces behind the rise in cohabitation. Intuitively, as wages grow and the price of home appliances decreases, household can buy on the the market goods that were once produced at home. Since one of the gain of marriage with respect to cohabitation lies in the better specialization of production within the household, we might expect that those forces imply a reduction in the share of couples choosing marriage over cohabitation.

The paper is organized as follows: section 2 offers an overview of US divorce laws, section 3 presents the empirical results, while section 4 describes in detail the theoretical model. Section 5 explain the procedure used for the estimation as well as the results. Section 6 report the results from a series of counter-factual experiments, while section 7 performs a welfare analysis. Section 8 draws the conclusions of the paper.

\section{US Divorce Laws: an Overview}
\section{Data and Empirical Evidence}
\subsection{The Dataset}
We begin by describing our data. We use the the National Survey of Family and the Household (NSFH) wave I, and the National survey of Family Growth, 1988 wave. Both surveys were designed to study the causes and consequences of changes happening in families and households within the United States. This is reflected in detailed question regarding the retrospective family history of respondents, which includes information both about marriage and cohabitation. Moreover, primary respondents\footnote{One adult per household was randomly selected as the primary respondent, while in the NSFG respondents are all women of 15-44 years of age.} are asked a large set of questions regarding their socio economic background and the demographics of the household. While the NSFH I is the first of three longitudinal waves, NSFG is made of several repeated cross sectional samples. We decided not to use the other two other waves of the NSFH because in the second wave all currently cohabiting households were dropped from the survey, while the 1988 wave of NSFG is the only one with publicly available information about the residence of the respondents. A drawback of using this data\footnote{We do not have the choice of using other surveys for our analysis, since they either lack the State of residence variable, or they miss information about cohabitation history, or they do not cover people that were in a relationship age at the time the law changed.} is that we know the state of residence of the respondents only at age 16 for the NSFH and at birth for the NSFG. Since we also know whether people lived all their life in the same State we are able to perform our empirical analysis both on the universe and of the subsample of never movers. We will show that point estimates turn out to be statistically not distinguishable between the two samples. Further details regarding those two surveys can be found in \cite{bumpass2017} and \cite{mosher1996}. We use this dataset to build two samples, the one of \textit{first and second relationships} and the one of \textit{first cohabitations}, that are described below.  
\subsubsection*{First and Second Relationships Sample}
We want to build a sample to analyze the type of relationship, which can be either marriage or cohabitation,\footnote{Dating is not considered, since we cannot observe this state. Hence, people dating will fall under the category of singles.} that respondents decided to have. The sample is made of first and second relationships. One first relationship is defined observing the first time (if ever) a certain person started cohabiting or married. This observation is associated to the date at which the relationship starts, to the characteristics of the respondent member of the formed couple, and with a \textit{type}, which can either marriage or cohabitation. Note the type of couples that cohabited before marriage is "cohabitation": the transition from cohabitation to marriage is analyzed using the sample of \textit{first cohabitations}. Second relationships  are defined in a similar fashion, but they include only respondents that ended the relationship with their first partner and started a new one with a different person. The way this sample is built implies that for some respondents we will have zero corresponding observations in this sample, while for other we will have other, and for others we will have two. We did not consider third or higher order relationships for our analysis since these individuals would be further away from the age at which we knew their state of residence. In [FIGURE********] we report the descriptive statistics of this sample.
\subsubsection*{First Cohabitation Sample}
This sample is built to analyze the decisions of cohabiting couples to separate or to marry. It is composed of the first non marital cohabitation experienced by respondents. This sample includes couples that cohabited before marriage, but it also includes cohabitations experienced by people with the following marital history: marriage without premarital cohabitation, divorce, cohabitation with a different person. Each observation of this sample is associated with a starting date, a possible ending date, and an outcome, which can be: still cohabiting, married or separated. In [FIGURE********] we report the descriptive statistics of this sample.
\subsection{Empirical Evidence}
\section{The Model}
This section describes a dynamic life cycle structural model of partnership formation, savings, female labor supply and  home production. Couples act cooperatively subject to limited commitment, which means that rebargaining might happen in response to changes in the outside options, which are assumed to be divorce or separation.

>>>>

\documentclass[12pt]{article}
%\documentclass[AEJ]{AEA}
\usepackage[round, sort , authoryear]{natbib}

\usepackage{pgf}
%%% PAGE DIMENSIONS
\usepackage[margin=2cm]{geometry}
%\usepackage[top=1.5in, bottom=1.5in, left=1.5in, right=1.5in]{geometry}
\usepackage{blindtext} % to change the page dimensions
%\geometry{a4paper} % or letterpaper (US) or a5paper or....
%\geometry{margin=1.5in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

@@

%Write a vecor
\newcommand{\myvec}[1]{\ensuremath{\begin{pmatrix}#1\end{pmatrix}}}

@@

%%% PAGE DIMENSIONS
%\usepackage[margin=2.5 cm]{geometry}
%\usepackage{blindtext} % to change the page dimensions
%\geometry{a4paper} % or letterpaper (US) or a5paper or....
%\geometry{margin=1.5cm} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage[utf8]{inputenc} 
\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{epstopdf}
\usepackage[hang]{footmisc}
\usepackage{lipsum}
\usepackage{setspace}
% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{pgfplots}
\pgfplotsset{compat=1.13}
\usepackage{caption}
\usepackage{threeparttablex}
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}
%%% PACKAGES
\usepackage{placeins}
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim

% These packages are all incorporated in the memoir class to one degree or another...
%\usepackage{amsmath}
\numberwithin{table}{section}
\usepackage{cases}
\usepackage{graphicx}
%
\usepackage{float}
\usepackage{authblk}
\usepackage{pgfplots}
\usepackage{pdfpages}
\linespread{1.5}
\setlength{\footnotemargin}{4mm}
\usepackage{amssymb} 
\usepackage{tabularx}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\addtolength{\footnotesep}{2mm} % change to 1mm

@@

%No Counting within section
\DeclareCaptionLabelSeparator{none}{}
\captionsetup{labelsep=none}
\counterwithout{figure}{section}
\counterwithout{table}{section}

%footnote space
\setlength{\footnotesep}{0.35cm}
\interfootnotelinepenalty=10000

%Nice Figure and table headers
\captionsetup[figure]{labelfont={sc},name={Figure},labelsep=period}
\captionsetup[table]{labelfont={sc},name={Table},labelsep=period,justification=centering}

%For subtables
\usepackage{subcaption}
\usepackage[justification=centering]{caption}
%For having the catption above tables
\usepackage{float}
\floatstyle{plaintop}
\restylefloat{table}

%For bar charts in table
\newlength\MAX  \setlength\MAX{20mm}
\newcommand*\Chart[1]{#1~\rlap{\textcolor{blue!20}{\rule{#1\MAX}{2ex}}}}
%\newcommand*\Chartguys[1]{\rlap{\textcolor{blue!30}{\rule{6ex}{\MAX}}}{\textcolor{white!30}{\rule{6ex}{#1\MAX}}}}
%\newcommand*\Chartgirls[1]{\rlap{\textcolor{red!30}{\rule{6ex}{\MAX}}}{\textcolor{white!30}{\rule{6ex}{#1\MAX}}}}
\newcommand*\Chartguys[1]{{\textcolor{blue!30}{\rule{6ex}{#1\MAX}}}}
\newcommand*\Chartgirls[1]{{\textcolor{red!30}{\rule{6ex}{#1\MAX}}}}

\renewcommand{\thesubfigure}{ (\alph{subfigure})}
\captionsetup[sub]{labelformat=simple}
%Math stuff
\DeclareMathOperator*{\argmax}{arg\,max}

@@

%\restylefloat{table}
%\oddsidemargin  +0.18in
%\evensidemargin +0.18in
%\topmargin 40pt \textheight 8.1in \textwidth 6.5in
%\linespread{1.5}\parskip .05in

%Nice Figure and table headers
\captionsetup[figure]{labelfont={sc},name={Figure},labelsep=none}
\captionsetup[table]{labelfont={sc},name={Table},labelsep=none,justification=centering}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\parskip3mm\parindent0cm

\title{The Rise of Cohabitation and Unilateral Divorce\thanks{The authors wish to thank David de la Croix, Matthias Doepke, Fabio Mariani, Jesús Fernández-Villaverde, Alessandra Voena and the Norhtwestern Macro group for their useful comments. Computational resources were provided by the supercomputing facilities of the Université catholique de Louvain (CISM/UCL) and Northwestern University. Blasutto acknowledges financial support from the French speaking community of Belgium (ARC project 15/19-063 on \textit{family transformations} and \textit{mandat d' aspirant} FC 23613).}}
\author{\large Fabio Blasutto$^1$ \quad Egor Kozlov$^2$}
\date{}

@@

\vspace{-2cm}
\begin{center}\large	
	Job Market Paper\\
	\href{https://drive.google.com/file/d/1utFkMPdUR5yb07oyJI7Vx1BJDLfYgMHi/view?usp=sharing}{Latest version available here}
\end{center}
\begin{center}\large
	\today
\end{center}

	\footnotetext[1]{IRES/LIDAM, UCLouvain \& FNRS (Belgium). Email:  \tt{fabio.blasutto@uclouvain.be}}
	\footnotetext[2]{Northwestern Univrsity. Email:  \tt{egorkozlov2020@u.northwestern.edu}.\\}

%\begin{abstract}
%	Marriage differs from informal cohabitation for its higher cost of breaking up. Moving from a mutual to a unilateral divorce regime drastically increases the risk of dissolution. We explore how this change affects partnership choice and the relative power within the couple. Exploiting the staggered introduction of unilateral divorce across the US states, we show that after the reform newly formed relationships are significantly more likely to be cohabitations instead of marriages, and that cohabitation spells last longer. To understand the drivers of these changes, we build and estimate a structural life cycle model with partnership choice, female labor supply and savings decisions, where the gains of marriage with respect to cohabitation come from a better risk sharing and specialization within the household, enforced through a costly divorce. In the model, the reform increases the risk of divorce, making cohabitation preferred by couples that would have had the highest risk of divorce under the mutual consent regime. Since the match quality of cohabitations is on average lower than that of marriages, the reform increases their length: we estimate that time spent cohabiting would have been four times lower if the law had never changed. In states where assets upon divorce are split equally, men would lose most of their wealth upon divorce: at equilibrium, many women accept to cohabit in exchange for a higher bargaining power. 
%\end{abstract}

%\begin{abstract}
%	This paper analyzes the role of unilateral divorce for the rise in cohabitation. Exploiting the staggered introduction of unilateral divorce across the US states, we show that after the reform singles become more likely to cohabit than to marry, and that newly formed cohabitation spells last longer. We then provide a theoretical rationale for these facts, building a life-cycle model with partnership choice, female labor supply and saving decisions. A structural estimation of the model suggests that the decline in marriage is due to an increased risk of divorce, while new cohabitation spells last longer for a selection effect. A counterfactual experiment reveals that time spent cohabiting would have been four times lower if divorce law had never changed.
%\end{abstract}

%\begin{abstract}
%	This paper analyzes the role of unilateral divorce for the rise of cohabitation. Exploiting the staggered introduction of unilateral divorce across the US states, we show that after the reform singles become more likely to cohabit than to marry, and that newly formed cohabitation spells last longer. We then provide a theoretical rationale for these facts, building a life-cycle model with partnership choice, female labor supply and saving decisions. A structural estimation of the model suggests that unilateral divorce reduces the gains from marriage, making cohabitation preferred to couples that would have had the highest risk of divorce. Since married couples are on average better matched, the reform increases the length of newly formed cohabitations because of a selection effect. A counterfactual experiment reveals that time spent cohabiting would have been four times lower if divorce law had never changed.
%\end{abstract}

\begin{abstract}
	This paper analyzes the role of unilateral divorce for the rise of cohabitation. Exploiting the staggered introduction of unilateral divorce across the US states, we show that after the reform singles become more likely to cohabit than to marry, and that newly formed cohabitations last longer. We then provide a theoretical rationale for these facts, building a life-cycle model with partnership choice, female labor force participation and saving decisions. A structural estimation of the model suggests that unilateral divorce decreases couples' stability, which makes cohabitation preferred to couples that would have been at high risk of divorce. Since cohabiting couples formed after the reform are better matched, the average length of cohabitations increases. A counterfactual experiment reveals that the time spent cohabiting would have been halved if divorce laws had never changed.
\end{abstract}

%\clearpage
%\noindent \medskip{}
\clearpage
\section{Introduction}

%Informal cohabitation is on the rise: the share of women that ever cohabited in the United States moved from 33\% in 1987 to 60\% in 2010 \citep{manning2013}. Since cohabiting couples display different outcomes than married ones, this rise definitely contributes to the overall changes in the structure and behavior of the American family. For example, \cite{bumpass2000} claim that the higher instability of cohabitation is responsible for the rise in the number of single mothers, which is linked to worse children's outcomes \citep{chetty2018}. The first step to find out why cohabitation is surging and why its outcomes differs from marriage is to better understand partnership choice, selection and cohabiting couples behavior.

%Moreover, \cite{poortman2012} point out that couple specific investments of those who cohabit are lower than those who are married.
%Since disadvantaged social classes are those who are more likely to cohabit instead of marrying, partnership choice can be an engine of economic inequality

Informal cohabitation is on the rise: the share of women that ever cohabited in the United States moved from 33\% in 1987 to 60\% in 2010 \citep{manning2013}. This increase contributed to the overall changes in the structure and behavior of the American family. For example, \cite{bumpass2000} claim that the higher instability of cohabitation is responsible for the rise in the number of single mothers, which is associated with poor outcomes of children \citep{chetty2018,mclanahan2013}. Moreover, compared to married couples, the relationship-specific investments of cohabitants are lower \citep{poortman2012} and their children's well-being is worse \citep{brown2004}, even after controlling for parental resources. However, it is not clear to which extent the outcomes of cohabiting families are due to selection versus the direct effect of partnership's rights on couple's behavior. Quantifying the relative importance of these two mechanisms can be done only by understanding the rational for partnership's choices. Why do people cohabit instead of marrying? Why did cohabitation become more common over time?

This paper addresses these questions by asking what role does unilateral divorce have for the rise of cohabitation. Since marriage and cohabitation can be viewed as contracts whose attractiveness depend on their termination rights and costs, the switch from mutual consent to unilateral divorce represents a unique opportunity to learn about partnership choices. This paper answers to this question with four contributions. First, we offer empirical evidence that unilateral divorce made cohabitation more likely to be chosen and that newly formed cohabitations last longer. Second, we propose a theory of partnership choice and endogenous breakup/divorce. In particular, in our model the policy makes cohabitation the preferred choice to couples that would have experienced the highest risk of a costly divorce. Therefore, the average match quality of cohabiting couples increases, which cause cohabitation spells to last longer. Third we structurally estimate our model to reflect the empirical findings. Fourth, we perform a counterfactual experiment to assess the overall role of unilateral divorce on the rise of cohabitation and we quantify the welfare gains and losses of the reform for men and women. This paper thus consists of four parts, one for each contribution, which we now describe in more detail.

%Since the seminal work of \cite{becker1981}, economists studied the incentives behind the decision to marry, among which the sharing of public goods, the division of labor to exploit comparative advantages (\citealp{chiappori1997}), and risk sharing (\citealp{voena2015} and \citealp{rigas2015}). These reasons tell us why couples live together, while they are actually silent about the choice between marriage and informal cohabitation. Those two partnerships can be viewed as contracts subject to different rules, which in particular regulate costs and rights to end these agreements. While the cost of divorce is higher than the one of just breaking up,\footnote{Henceforth we will refer to the separation from cohabitation as breakup to avoid confusion with the legal separation.}, the introduction of unilateral divorce during the 1970s shifted the property rights from the spouse who wished to stay married to the one who wished to divorce. Hence, the switch from mutual consent to unilateral divorce represents a unique opportunity to learn how people choose between marriage and cohabitation.
%In particular, unilateral divorce might have eroded the gains from marriage, both increasing the likelihood of a costly divorce and hampering couples' ability to commit, thus making some couples willing to cohabit instead of marrying. In addition, shifting to a less committed relationship hampers couples' ability to specialize and it might affect the relative bargaining  power of partners.

%In this paper, we study the role of unilateral divorce on the rise of cohabitation. Our main contribution is to show that the rate of cohabitation surged after the changes in the US law because the increased risk of divorce made marriage less attractive. In particular, cohabitation becomes the preferred choice to couples that would have experienced the highest risk of divorce. As a consequence, the average match quality of cohabiting couples increases, which cause cohabitation spells to last longer. Interestingly, men are those who wish to cohabit the most after the reform because they would lose a relatively higher share of their wealth upon divorce. This increases the bargaining power of women entering cohabitation.

In the first part of the paper we document the effect of unilateral divorce on the choice between marriage and cohabitation and on the duration of newly formed cohabitations. We use data from the National Survey of Family and the Household (NSFH) and from the National Survey of Family Growth (NSFG) to study the choice between marrying and cohabiting. Then, exploiting the exogenous variation coming from the staggered introduction of unilateral divorce over time across the US states, we estimate that newly formed couples are 5-7\% more likely to choose cohabitation over marriage compared to the pre-reform period. Interestingly, the effect's size depends on how property is divided upon divorce, being strongest in states where each spouse gets half of the wealth and where the assets shares are decided by the judge. This suggest that divorce allocations affect partnership choices.
%This suggests that the result cannot be entirely driven by divorce being a less binding commitment technology,\footnote{This mechanism is highlighted by \cite{lafortune2020}, where the equal division of assets within marriage makes it possible to efficiently invest in children even after the rise of unilateral divorce.} since the egalitarian property division rule can act as a substitute for the mutual consent regime.
Moreover, we analyze how unilateral divorce affected the duration of cohabitation spells: our estimates from a multinomial probit show that cohabitations formed after the reform last longer both because of a reduced risk of marriage and of breakup.\footnote{Henceforth we will refer to the separation from cohabitation as breakup to avoid confusion with legal separation.}

In the second part of the paper we propose a theory to understand the mechanisms underlying the facts documented in the empirical part. We build a dynamic model of intra-household decision making and search in the mating market, where agents makes decisions according to the realization of idiosyncratic income shocks, their amount of wealth and a couple-specific match quality. Single agents meet, with some probability, a potential partner with a match quality, productivity and wealth. After the draw, they decide whether to marry, cohabit or stay single. Couples make decisions about consumption, savings and female labor force participation. Women receive a productivity penalty for not working and women's time can be used to produce a public good which captures utility gains from children, durable goods and services. Cohabitation and marriage differ in their splitting costs as well as in the way property is divided when the couple dissolves. Moreover, cohabitation is subjected to stigma, which is modeled as an exogenous utility flow.\footnote{The stigma towards cohabitation is to be interpreted considering the shame towards out-out-wedlock births, premarital sex and premarital cohabitation at time of the divorce revolution.} In the case of breakup, assets are split according to individual property rights, while in the case of divorce they are divided in half.\footnote{We made this assumption because the model will be estimated using community property states data.} Breakup (unilateral divorce) can be initiated unilaterally, as opposed to mutual consent divorce, which requires the consensus of both partners. Following \cite{voena2015}, under mutual consent the couple always cooperates while married, which implies that the allocation of resources corresponds to the Pareto-efficient inter-temporal allocation. Instead, unilateral divorce (breakup) results in a lack of commitment within marriage (cohabitation),\footnote{We model the decision making in the couple building on the literature on limited commitment (see for example \cite{kk1996},  \cite{ligon2002}, \cite{marcet2019} and \cite{pavoni2018}), which has been applied to dynamic collective models in the household by \cite{mazzocco2007}, \cite{mazzocco2013}, \cite{bayot2015}, \cite{rigas2015}, \cite{voena2015}, \cite{abraham2018}, \cite{lise2018}, \cite{low2018}, \cite{foerster2019} and \cite{reynoso2019} among others.} in fact the intra-household allocation can change in response to shocks, as spouses (cohabitants) can credibly exercise the threat of divorce (breakup). The fact that couples cannot commit ex-ante to any possible division of marital surplus contributes to the creation of an imperfectly transferable utility environment, under which the Becker-Coase theorem does not hold.\footnote{For an application of an imperfectly transferable utility environment to matching and collective models see \cite{galichon2019} and \cite{weber2018}.} Hence, abandoning the mutual consent regime affects the risk of divorce and in turn the surplus of marriage.\footnote{According to \cite{becker1977}, divorce laws should not affect separation decisions \textit{``if all compensations between spouses were feasible and costless"}. The assumptions underlying the Becker-Coase theorem are discussed by \cite{chiappori2015} and \cite{fella2004}.} More precisely, the barriers to divorce---represented by its costs and rights---affect the gains of marriage with respect to cohabitation through three main channels. First, by acting as commitment technologies, they enforce a better risk sharing and a more efficient specialization in the production of public goods within marriage.\footnote{We assume that public goods are produced with money and women's time. When women stop working for producing such a good their potential productivity in the labor market decreases.} Second, they increase the risk of being ``trapped" in a bad marriage. Third, they affect the expected value of marriage by modifying the risk and value of divorce. The effects of tightening or relaxing the barriers to divorce depends on which channel prevails: for example, the introduction of unilateral divorce has an uncertain impact on the share of couples that cohabit and marry. 
Interestingly, in our framework  the outcomes of cohabitation and marriage not only depend on the rules underlying these contracts, but also on sorting on match quality. In fact, a cheap breakup is most attractive to couples whose match quality is low and couples' ability to cooperate depends on the match quality itself through its effect on the couples' stability.\footnote{It's easier to make partnership-specific investments when the risk of splitting is low, which is the case when the match quality is high.}

%
%In our framework marriage gains depend on the quality of the match: for couples whose value is low cohabitation is the best choice because breaking up is cheaper and their risk of splitting is high. Interestingly, the ability of the couple to cooperate is directly affected by the quality of the match. For this reason, the outcomes of cohabitation and marriage depend not only of the rules underlying these contracts, but the ability of the couple to cooperate is directly affected by the quality of the match.
